JD notes on effects

### 2022-06-13 manuscript/body.tex
All chapters draft
Draft sync

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}

@@

In applications that use regression models, outcome predictions are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with\bmb{?only}\BC{clarify with BMB} interactions, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.

Ordinary least squares (OLS) regression provides an easy way to express multivariate relationships as an additive linear combination of many predictors with the outcome. The coefficient estimates from OLS regression models are generally easy to interpret as the expected change in the outcome with a unit change in the continuous predictor or as the expected outcome difference between the levels of the categorical predictor with the baseline category. However, with additional complexities such interaction between predictors, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.

Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including \bmb{nonlinear relationships -- need to clarify}\BC{clarity with BMB} between response and predictors e.g., log transformed response or predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions \bmb{due to their flexibility}\BC{clarify why BMB circled it}. This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}. Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model. An alternative is to explore the \emph{effects} of predictors on the predictions of the outcome at various levels of the predictor of interest \jd{Are we still saying these (prediction and effects) are the same? Or are we now saying they're different, with the same citations?} \citep{fox2009effect, leeper2017package, lenth2018package}. For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.

When visually presented, predictions provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret. Generating predictions together with the associated confidence intervals for regression models has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of \emph{focal} predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
\item  propagation of uncertainty -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles \jd{Don't understand what the bins are doing here (when we are holding the reference point constant); it's not what we do, is it?} if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{hanmer2013behind}. This procedure generates \jd{Are predictions and predictor effects really used for the same thing?} -- \emph{predictions}, \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as predictions since they should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''. \bmb{More specifically, we compute the expected outcome by holding non-focal predictors constant (or averaged in some meaningful way) while varying the focal predictor, with the goal that the predictions represents how the model responds to the changes in the focal predictor. On the other hand, effects differ with predictions in the sense that the former is about how we describe the uncertainty around the latter and depends on the anchor. -- see next/unclear}\BC{clarify with BMB}

The most \bmb{commonly used \proglang{R} software packages (\pkg{emmeans} and \pkg{effects})}\BC{not sure the underline} for generating predictions, by default, use the average of non-focal predictors as the reference point. 
\jd{This is ambiguous; we need to be talking about input variables and model variables by now. Also: what do the packages do about categorical non-focal predictors? Do we know?}
However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting model variables) versus taking the product of the averages of the interacting model variables (the default for \pkg{emmeans} and \pkg{effects}). For simple models (with no interactions), the two approaches give the same predictions. However, for complex models, we claim that neither of these two approaches is the most appropriate but we show that averaging the interaction closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
\bmb{This whole example .... example of nonlinear averaging}\BC{Yes and handling interaction terms?}
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
\end{align}
%

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
\jd{What are true predictions? Why does it look like effects matches varpred, but it sounds like we're saying it matches emmeans?}

@@

\section*{Quantities of interest}

@@

\subsection*{Input variables}

@@

\subsection*{Model matrix}

@@

\subsection*{Model variables}

@@

\subsection*{Model center}

@@

\subsection*{Reference point}

@@

\subsection*{Anchor}

@@

\subsection*{Prediction- vs. effect- styled plots}

@@

\subsection*{Predictions vs. effects}

@@

\section*{Statistical background}

@@

\subsection*{Dealing with higher order interactions}

@@

\section*{Uncertainty estimation}

@@

\subsection*{Model center based}

@@

%% \subsection*{Is center-anchored better?}
%% 
%% We want to compare the amount of uncertainty when there is/no anchor. Let $\bao = \boldsymbol{0}$ be the anchoring matrix with all entries in $0$s, and let $\bafc$ be the center-anchored matrix defined above. Then entries in $(\bxo - \bao) \geq (\bxo - \bafc)$.
%%

\section*{Bias correction}

@@

\subsection*{Whole-sample-based approach for bias correction}

@@

\section*{Mean-based vs. whole-sample-based}

@@

\section*{Simulation examples}

@@

\subsection*{Predictions-styled and effects-styled plots}

@@

\subsection*{Bias correction}

@@

\subsection*{Mediated effect}

@@

\section*{Discussion}

@@

\section*{Conclusion}

@@

\section*{Supporting information}

>>>>

% Use "Eq" instead of "Equation" for equation citations.
\section{Introduction}

@@

In applications that use regression models, outcome predictions are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.

Ordinary least squares (OLS) regression provides an easy way to express multivariate relationships as an additive linear combination of many predictors with the outcome. The coefficient estimates from OLS regression models are generally easy to interpret as the expected change in the outcome with a unit change in the continuous predictor or as the expected outcome difference between the levels of the categorical predictor with the baseline category. However, with additional complexities such as interactions between predictors, the resulting coefficients risk misinterpretation or interpretation complexity \citep{brambor_understanding_2006, berry_improving_2012}.

Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and predictors e.g., nonlinear transformation on the predictors, interactions between predictors (and via splines for example), and nonlinear transformations via link functions. This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}. Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model. An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.  For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.

Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} predictor (predictor of interest). The major difference between the two lies in the construction of the associated confidence intervals. In particular, for predictions, we want to use the classic curved confidence intervals (or prediction intervals) which capture the uncertainty in the whole model and generally focus on the \emph{total} effects, and so will often be suited for a univariate model. On the other hand, effects, attempt to show the effect of a focal predictor. In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal predictor only, and so will often be suited for the direct effects – after controlling for other predictors – from a multivariate model.

When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the predictors whose estimates are usually, but not always, harder to interpret. Generating these quantities together with the associated confidence intervals for regression models has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of focal predictor and the \emph{reference point} for non-focal predictors, especially in multivariate models
\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal predictors 
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

The most common way of choosing the representative values is taking unique levels of the focal predictor if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other predictors other than the focal predictor) predictors at their reference point (e.g., means) \citep{fox2009effect, hanmer2013behind}. This procedure generates -- predictions (at least in our case), \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the predictor, for an ``average case''. More specifically, the basic idea is to compute the expected outcome by holding the non-focal predictors constant, for example at their mean (centered), while varying the focal predictor, with the goal that estimates represent how the model responds to the changes in the focal predictor. As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.

The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of input and model variables as the reference point for continuous and categorical predictors, respectively. However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting model variables) versus taking the product of the averages of the interacting model variables (the default for \pkg{emmeans} and \pkg{effects}). For simple models (with no interactions), the two approaches give the same predictions. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions (model variables) closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal predictor:
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}
\end{align}
%
We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.

@@

\section{Quantities of interest}

@@

\subsection{Input variables}

@@

\subsection{Model matrix}

@@

\subsection{Model variables}

@@

\subsection{Model center}

@@

\subsection{Reference point}

@@

\subsection{Anchor}

@@

\subsection{Prediction- vs. effect- styled plots}

@@

\subsection{Predictions vs. effects}

@@

\section{Statistical background}

@@

\subsection{Dealing with higher order interactions}

@@

\section{Uncertainty estimation}

@@

\subsection{Model center based}

@@

%% \subsection{Is center-anchored better?}
%% 
%% We want to compare the amount of uncertainty when there is/no anchor. Let $\bao = \boldsymbol{0}$ be the anchoring matrix with all entries in $0$s, and let $\bafc$ be the center-anchored matrix defined above. Then entries in $(\bxo - \bao) \geq (\bxo - \bafc)$.
%%

\section{Bias correction}

@@

\subsection{Whole-sample-based approach for bias correction}

@@

\section{Mean-based vs. whole-sample-based}

@@

\section{Simulation examples}

@@

\subsection{Predictions-styled and effects-styled plots}

@@

\subsection{Bias correction}

@@

\subsection{Mediated effect}

@@

\section{Discussion}

@@

\section{Conclusion}

@@

\section{Supporting information}
